Comparing the Jacobi Method and LLL lattice reduction algorithms for cryptographic applications

### 2014-12-22 report/paper.tex
Review plus adding running time graph for dimension increase

\title{Comparing the Jacobi Method and LLL lattice reduction algorithm\my{s} for cryptographic applications}
\date{Fall 2014}
\author{Frederic Jacobs\\ EPFL Bachelor Semester Project\\ frederic.jacobs@epfl.ch}

@@

\section{Introduction}
In 2012, Sanzheng Qiao published a paper~\cite{originalJacobiMethodLatticeBasisReduction} proposing the use of the Jacobi method for lattice reduction applications. In following papers, Qiao proposed several enhancements to the generic Jacobi method. \my{The purpose of their research is to generate a reduction that produces good reduction on random matrices for which the columns represent almost orthogonal vectors. This is in contrast to lattices that are used in cryptography as public keys. From the cryptanalytic point of view, we are interested in random lattices where the given basis vectors are skewed and not almost orthogonal.} We implemented the Jacobi method reduction algorithm \my{as presented by Qiao et al. and compared the quality of the outputs with the LLL algorithm for almost orthogonal matrices and random lattices. The results show...}

@@

The Euclidean norm in $\mathbb{R}^n$ is a mapping from $\mathbb{R}^n $ \my{to} $\mathbb{R}$ that assigns \my{a} length to any vector of the space \my{$\mathbb{R}^n$}.

@@

\caption{\my{text text text}}
\label{fig:latti}
\end{figure}

This can be translated into a more formal definition~\cite{SchnorrStanfordNotes} \my{as follows}.\newline 
Let $\vec{b_1}, \vec{b_2}, ... \vec{b_n} \in \mathbb{R}^m$ be linearly independent vectors for $n\leq m$. We call the discrete additive subgroup

@@

A lattice $L \subseteq \mathbb{R}^m$ is \emph{full dimensional} when $rank(L)=m$. We will work exclusively with full dimensional lattices. This means that we will work on matrices with $m=n$ linearly independent vectors \my{in $\mathbb{R}^n$}.

@@

As this article focuses on lattice basis reduction, it is important to define the equivalence of two bases. Two bases $B_1 , B_2$ are equivalent iff $B_2 = B_1 U$ for a unimodular matrix $U$.
A \emph{unimodular matrix} $U$ is a square integer matrix such that $det(U) = \pm 1$. \my{$B_1$ and $B_2$ are bases to the same lattice.} The right multiplication of a matrix $B_1$ with a unimodular matrix $U$ yields $B_2$ by performing a combination of the following \my{elementary integer} column operations:
\begin{itemize}
\item Swap two columns
\item Negate the elements of a column
\item Addition of an integer multiple of one column to another
\end{itemize}

@@

Without loss of generality, we hence restrict tests to random lattices of volume $P$ whose basis in HNF form is as follows: \my{no bold ai's in the matrix}
$$\left[\begin{array}{ccccc}
P & \vec{a_2} & \dots & \vec{a_{m}}\\
& 1&   & \\ 
& & \ddots & \\ 
 & & &1 
\end{array}\right]$$
where $|a_i| < P$.

@@

The orthogonality defect is a quality indicator that we will use to compare the reduction of \emph{the set of basis vectors}. As the basis vectors are getting \my{increasingly? more} orthogonal during a reduction, the nominator will become smaller. The lower bound of the orthogonality defect is $\text{OrthDefect} \ge 1$. In the special case where the basis would be completely orthogonal, the orthogonality defect would be $\text{OrthDefect} = 1$.

@@

Given a basis of a\my{n} $n$-rank integral lattice $L$, find $\vec{u} \in L$ such that $\|\vec{u}\| = \lambda_1 (L)$

@@

\subsubsection{Minkowski's Convex Body Theorem}
\my{Still check grammar in following sentence.}
For a full-rank lattice $L$ of $\mathbb{R}^n$ and $C$, a measurable subset of $\mathbb{R}^n$, convex, symmetric with respect to 0, and of measure $> 2^n \text{vol}(L)$. Then $C$ contains at least a nonzero point of the lattice $L$.

A corollary of Minkowski's theorem tells us that there exists for any $n$-dimensional full-rank lattice $L$ a nonzero lattice-vector $x$ whose length is bounded by $\sqrt{n} (\det L)^{1/n}$. So we can derive a bound on the length of a shortest lattice-vector:
\[
\lambda_1(L) \leq \sqrt{n} (\det L)^{1/n}.
\]
\my{In the 80's lattice reduction showed to be successful in finding shortest vectors whose length was significantly smaller than indicated by the above bound.}

@@

The GSO process  can be defined recursively:
\[
\begin{cases}
\vec{b_1}^{\star} = \vec{b_i}  \\
\vec{b_i}^{\star} = \displaystyle\sum^{i-1}_{j=1} \mu_{i,j} \vec{b_j^\star}
\end{cases}
\]
where $\mu_{ij} = \frac{\langle \vec{b}_i, \vec{b}_j^\star \rangle}{\| b_j^\star\|^2}$. The set of vectors $b_i^*$ are only constituting a basis of the lattice if their $b_i^* \in \mathbb{Z}$ \my{. No, if $\mu $'s are integers.}

@@

Since the LLL-reduction algorithm returns a $\delta$-LLL reduced basis, we can observe the following properties on the basis:
\begin{itemize}
\item $\| \vec{b_1} \| \leq \alpha^{(n-1)/2} \lambda_1 $
\item $\max_i \| \vec{b_i} \| \leq \alpha^{(n-1)/2} \lambda_n $
\item $\| \vec{b}_1 \| \leq \alpha^{(n-1)/4} \det B^{1/n}$ \my{det(B) or det(L) ?  }
\end{itemize} 
where the parameter $\alpha = 1/(\delta - \frac{1}{4})$. Since $\delta$ is bounded by $1/4 < \delta < 1$, parameter $\alpha > 4/3$.

@@

The LLL algorithm, like many reduction algorithms, can be broken down into two parts: size-reduction and vector swapping. Size-reduction reduces the vectors from the input basis. After that, the vectors of the basis are checked against the Lovász condition. If a pair of vectors of the basis $(\vec{b_{i}}, \vec{b_{i+1}})$ don't \my{a pair is singular} satisfy the condition, we swap \my{the vectors} and go back to the size reduction (cfr. \textbf{Procedure \ref{alg:lllAlgorithm}}).

@@

Lagrange's algorithm solves the shortest vector problem exactly for lattice bases of dimension 2. It is a variant of Euler's centered algorithm that is used to compute the greatest common divider \my{still not divisor?} of two numbers. GCD can be thought to be a one dimension reduction problem. The GCD is the first minimum of the lattice $n \mathbb{Z} + m \mathbb{Z}$ spanned by $n$ and $m$.

\input{algorithms/lagrange}
\subsection{The Generic Jacobi Method}
We call \emph{generic Jacobi method} the algorithm that simply applies the Lagrange algorithm to every pair of the basis vectors for an $n$ dimensional lattice. This method was introducedby Qiao in 2012~\cite{originalJacobiMethodLatticeBasisReduction} claiming experimental results to be considerably faster than LLL \my{for almost orthogonal bases}.

@@

\subsubsection{$\omega$-Lagrange-reduced conditions}
There are two conditions for a basis to be $\omega$-Lagrange-reduced expressed with the Gram matrix $G$ of the basis $B$. The Gram matrix $G$ of a basis $B$ is computed \my{as} $G = B^T B$. By construction, a Gram matrix is always symmetric, for every pair of indices $(i,j)$, we have $g_{ij} = g_{ji}$. \my{A basis is $\omega$-Lagrange-reduced if it satifies: }

@@

\end{cases}
\]  where $1/\sqrt{3} \leq \omega < 1$ \my{and $1\leq s,\ell \leq n $}. The $s$ and $l$ indices are referring to the index of the short and long vectors in the basis Gram matrix of the basis.

\subsubsection{Size-reduction Lagrange}
The size-reduction algorithm \my{better LagrangeIT} is a variant of the Lagrange algorithm that we presented in section~\ref{sec:lagrangeAlgorithm}. The reduction is only performed if the $\omega$-Lagrange-reduced conditions are not satisfied. In which case, both the unimodular reduction matrix $Z$ and the Gram matrix $G$ are updated by applying the Lagrange reduction to the longest of the basis vectors at index $i$ or $j$.

@@

We implemented the Fast-Jacobi\my{~\cite{}} Jacobi reduction\my{, Proc.\ref{jacobiVariantReduction},} in C++ on top of the \emph{newNTL} library. The FPLLL implementation of LLL was used to compare the results after reduction.

@@

Our experimentation focused on the quality of the reduction. The two quality \my{metics?} we used for comparing the reductions are:

\begin{itemize}
\item the orthogonality defect (See section: \ref{sec:orthogonalityDefect}), gives us an indication about the size of the reduced basis vectors for a given lattice \my{in comparison to the volume}.
\item the Hermite factor (See section: \ref{sec:hermiteFactor}), gives us an indication about the size of the shortest vector of the reduced basis for a given lattice.
\end{itemize}

\subsection{Input bases}
Numerous papers were written about the Jacobi method for lattice reduction by S. Qiao , namely~\cite{originalJacobiMethodLatticeBasisReduction}, \cite{complexityAnalysisOfJacobiMethod} and ~\cite{enhancedJacobi}, but none of the papers \my{is} clarifying what bitsize was used for the randomly generated entries of the vectors. The main focus of the authors was the use of lattice reduction for signal processing, more specifically multiinput and multioutput (MIMO), the use of multiple antennas at both the transmitter and receiver to improve communication reliance. MIMO uses smaller dimensions lattices. In \cite{enhancedJacobi} and \cite{tian2014hybrid}, a reduction is performed on lattices of dimensions 20 to 100. These dimensions are inappropriate for cryptographic applications.

@@

\subsection{On Almost Orthogonal Basis \my{plural?}}

We initially ran our Jacobi reduction variant on almost orthogonal bases to attempt to reproduce Qiao's results. The orthogonality defects are similar to what the authors found in \cite{tian2014hybrid} and \my{\cite{fastJacobi}}. Our results are shown in Fig.~\ref{fig1} and Fig.~\ref{fig2}. By increasing the parameter $\omega$, we are making the $\omega$-Lagrange-reduced conditions more restrictive. More specifically, for every pair of vectors of the basis, the following condition needs to be satisfied
\[
 \omega \| \vec{a}_l \| \leq \|\vec{a}_l \pm \vec{a}_s \|
\] where $\vec{a}_l$ is the longest vector and $\vec{a}_s$ is the shortest. Hence, the reduction renders the basis more orthogonal as $\omega$ grows. \my{We see that for $\omega=0.6$ , LLL outperforms... while for $\omega = 0.9$...}

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.5\textwidth]{results-graphs/runningtime.png}

\caption{Running time increasing substantially with $\omega$ parameter}
\label{fig3}
\end{figure}

@@

We notice that the Jacobi algorithm does reduce the bases further than LLL. Looking at the conditions to be $\delta$-LLL reduced and $\omega-$Jacobi reduced, we can find for example two bases that are LLL reduced but not Jacobi reduced \my{and will miss the shortest vector}.

@@

Although the Jacobi method might have practical applications in signal processing, it is not appropriate for cryptographic applications on random lattices or higher dimension\my{al} bases.

>>>>

\title{Comparing the Jacobi Method and LLL lattice reduction algorithms for cryptographic applications}
\date{Fall 2014}
\author{Frederic Jacobs\\ EPFL Bachelor Semester Project\\ frederic.jacobs@epfl.ch}

@@

\section{Introduction}
In 2012, Sanzheng Qiao published a paper~\cite{originalJacobiMethodLatticeBasisReduction} proposing the use of the Jacobi method for lattice reduction applications. In following papers, Qiao proposed several enhancements to the generic Jacobi method. The purpose of their research is to generate a reduction that produces good reduction on random matrices for which the columns represent almost orthogonal vectors. This is in contrast to lattices that are used in cryptography as public keys. From the cryptanalytic point of view, we are interested in random lattices where the given basis vectors are skewed and not almost orthogonal. We implemented the Jacobi method reduction algorithm as presented by Qiao et al. and compared the quality of the outputs with the LLL algorithm for almost orthogonal matrices and random lattices.

@@

The Euclidean norm in $\mathbb{R}^n$ is a mapping from $\mathbb{R}^n $ to $\mathbb{R}$ that assigns a length to any vector of the space $\mathbb{R}^n$.

@@

\caption{Two lattices in two-dimensional space. Left: Parallelogrammic lattice Right: Square lattice}
%wording from wikipedia: http://en.wikipedia.org/wiki/Lattice_(group)
\label{fig:latti}
\end{figure}

This can be translated into a more formal definition~\cite{SchnorrStanfordNotes} as follows:\newline 
Let $\vec{b_1}, \vec{b_2}, ... \vec{b_n} \in \mathbb{R}^m$ be linearly independent vectors for $n\leq m$. We call the discrete additive subgroup

@@

A lattice $L \subseteq \mathbb{R}^m$ is \emph{full dimensional} when $rank(L)=m$. We will work exclusively with full dimensional lattices. This means that we will work on matrices with $m=n$ linearly independent vectors in $\mathbb{R}^n$.

@@

As this article focuses on lattice basis reduction, it is important to define the equivalence of two bases. Two bases $B_1 , B_2$ are equivalent iff $B_2 = B_1 U$ for a unimodular matrix $U$.
A \emph{unimodular matrix} $U$ is a square integer matrix such that $det(U) = \pm 1$. $B_1$ and $B_2$ are bases to the same lattice. The right multiplication of a matrix $B_1$ with a unimodular matrix $U$ yields $B_2$ by performing a combination of the following elementary integer column operations:
\begin{itemize}
\item Swap two columns
\item Negate the elements of a column
\item Addition of an integer multiple of one column to another
\end{itemize}

@@

Without loss of generality, we hence restrict tests to random lattices of volume $P$ whose basis in HNF form is as follows: $$\left[\begin{array}{ccccc}
P & a_2 & \dots & a_{m}\\
& 1&   & \\ 
& & \ddots & \\ 
 & & &1 
\end{array}\right]$$
where $|a_i| < P$.

@@

The orthogonality defect is a quality indicator that we will use to compare the reduction of \emph{the set of basis vectors}. As the basis vectors are getting more orthogonal during a reduction, the nominator will become smaller. The lower bound of the orthogonality defect is $\text{OrthDefect} \ge 1$. In the special case where the basis would be completely orthogonal, the orthogonality defect would be $\text{OrthDefect} = 1$.

@@

Given a basis of an $n$-rank integral lattice $L$, find $\vec{u} \in L$ such that $\|\vec{u}\| = \lambda_1 (L)$

@@

\subsubsection{Minkowski's Convex Body Theorem}
Let $L$ be a full-rank lattice of $\mathbb{R}^n$ and $C$, a measurable subset of $\mathbb{R}^n$, convex, symmetric with respect to 0, and of measure $> 2^n \text{vol}(L)$. Then $C$ contains at least a nonzero point of the lattice $L$.

A corollary of Minkowski's theorem tells us that there exists for any $n$-dimensional full-rank lattice $L$ a nonzero lattice-vector $x$ whose length is bounded by $\sqrt{n} (\det L)^{1/n}$. So we can derive a bound on the length of a shortest lattice-vector:
\[
\lambda_1(L) \leq \sqrt{n} (\det L)^{1/n}.
\]
In the 80's lattice reduction showed to be successful in finding shortest vectors whose length was significantly smaller than indicated by the above bound.

@@

The GSO process  can be defined recursively:
\[
\begin{cases}
\vec{b_1}^{\star} = \vec{b_i}  \\
\vec{b_i}^{\star} = \displaystyle\sum^{i-1}_{j=1} \mu_{i,j} \vec{b_j^\star}
\end{cases}
\]
where $\mu_{ij} = \frac{\langle \vec{b}_i, \vec{b}_j^\star \rangle}{\| b_j^\star\|^2}$. The set of vectors $b_i^*$ are only constituting a basis of the lattice if their $\mu$ factors $\in \mathbb{Z}$

@@

Since the LLL-reduction algorithm returns a $\delta$-LLL reduced basis, we can observe the following properties on the basis:
\begin{itemize}
\item $\| \vec{b_1} \| \leq \alpha^{(n-1)/2} \lambda_1 $
\item $\max_i \| \vec{b_i} \| \leq \alpha^{(n-1)/2} \lambda_n $
\item $\| \vec{b}_1 \| \leq \alpha^{(n-1)/4} \det L^{1/n}$\end{itemize} 
where the parameter $\alpha = 1/(\delta - \frac{1}{4})$. Since $\delta$ is bounded by $1/4 < \delta < 1$, parameter $\alpha > 4/3$.

@@

The LLL algorithm, like many reduction algorithms, can be broken down into two parts: size-reduction and vector swapping. Size-reduction reduces the vectors from the input basis. After that, the vectors of the basis are checked against the Lovász condition. If a pair of vectors of the basis $(\vec{b_{i}}, \vec{b_{i+1}})$ don't satisfy the condition, we swap the vectors and go back to the size reduction (cfr. \textbf{Procedure \ref{alg:lllAlgorithm}}).

@@

Lagrange's algorithm solves the shortest vector problem exactly for lattice bases of dimension 2. It is a variant of Euler's centered algorithm that is used to compute the greatest common divisor of two numbers. GCD can be thought to be a one dimension reduction problem. The GCD is the first minimum of the lattice $n \mathbb{Z} + m \mathbb{Z}$ spanned by $n$ and $m$.

\input{algorithms/lagrange}
\subsection{The Generic Jacobi Method}
We call \emph{generic Jacobi method} the algorithm that simply applies the Lagrange algorithm to every pair of the basis vectors for an $n$ dimensional lattice. This method was introduced by Qiao in 2012~\cite{originalJacobiMethodLatticeBasisReduction} claiming experimental results to be considerably faster than LLL for almost orthogonal bases.

@@

\subsubsection{$\omega$-Lagrange-reduced conditions}
There are two conditions for a basis to be $\omega$-Lagrange-reduced expressed with the Gram matrix $G$ of the basis $B$. The Gram matrix $G$ of a basis $B$ is computed as $G = B^T B$. By construction, a Gram matrix is always symmetric, for every pair of indices $(i,j)$, we have $g_{ij} = g_{ji}$. A basis is $\omega$-Lagrange-reduced if it satifies:

@@

\end{cases}
\]  where $1/\sqrt{3} \leq \omega < 1$ and $1\leq s,\ell \leq n $. The $s$ and $l$ indices are referring to the index of the short and long vectors in the basis Gram matrix of the basis.

\subsubsection{Size-reduction Lagrange}
The LagrangeIT algorithm is a variant of the Lagrange algorithm that we presented in section~\ref{sec:lagrangeAlgorithm}. The reduction is only performed if the $\omega$-Lagrange-reduced conditions are not satisfied. In which case, both the unimodular reduction matrix $Z$ and the Gram matrix $G$ are updated by applying the Lagrange reduction to the longest of the basis vectors at index $i$ or $j$.

@@

We implemented the Fast-Jacobi~\cite{fastJacobi} Jacobi reduction, Proc.\ref{jacobiVariantReduction}, in C++ on top of the \emph{newNTL} library. The FPLLL implementation of LLL was used to compare the results after reduction.

@@

Our experimentation focused on the quality of the reduction. The two quality metrics we used for comparing the reductions are:

\begin{itemize}
\item the orthogonality defect (See section: \ref{sec:orthogonalityDefect}), gives us an indication about the size of the reduced basis vectors for a given lattice in comparison to the volume.
\item the Hermite factor (See section: \ref{sec:hermiteFactor}), gives us an indication about the size of the shortest vector of the reduced basis for a given lattice.
\end{itemize}

\subsection{Input bases}
Numerous papers were written about the Jacobi method for lattice reduction by S. Qiao , namely~\cite{originalJacobiMethodLatticeBasisReduction}, \cite{complexityAnalysisOfJacobiMethod} and ~\cite{enhancedJacobi}, but none of the papers is clarifying what bitsize was used for the randomly generated entries of the vectors. The main focus of the authors was the use of lattice reduction for signal processing, more specifically multiinput and multioutput (MIMO), the use of multiple antennas at both the transmitter and receiver to improve communication reliance. MIMO uses smaller dimensions lattices. In \cite{enhancedJacobi} and \cite{tian2014hybrid}, a reduction is performed on lattices of dimensions 20 to 100. These dimensions are inappropriate for cryptographic applications.

@@

\subsection{On Almost Orthogonal Bases}

We initially ran our Jacobi reduction variant on almost orthogonal bases to attempt to reproduce Qiao's results. The orthogonality defects are similar to what the authors found in \cite{tian2014hybrid} and \cite{fastJacobi}. Our results are shown in Fig.~\ref{fig1} and Fig.~\ref{fig2}. By increasing the parameter $\omega$, we are making the $\omega$-Lagrange-reduced conditions more restrictive. More specifically, for every pair of vectors of the basis, the following condition needs to be satisfied
\[
 \omega \| \vec{a}_l \| \leq \|\vec{a}_l \pm \vec{a}_s \|
\] where $\vec{a}_l$ is the longest vector and $\vec{a}_s$ is the shortest. Hence, the reduction renders the basis more orthogonal as $\omega$ grows. We see that for $\omega=0.6$ , LLL outperforms the Jacobi method while for $\omega = 0.99$, the Jacobi method outperforms LLL in terms of orthogonality defect.

@@

\begin{figure}[H]
\begin{minipage}{.45\textwidth}
\includegraphics[width=0.9\textwidth]{results-graphs/runningtime.png}

\caption{Running time (microseconds) increasing substantially with $\omega$ parameter}
\label{fig3}
\end{minipage}
\hspace{.05\textwidth}
\begin{minipage}{.45\textwidth}
    \includegraphics[width=0.9\textwidth]{results-graphs/runningtime-dimension.png}    
  \caption{Running time (microseconds) increasing substantially with dimension}
  \label{fig6}
\end{minipage}

\end{figure}

@@

We notice that the Jacobi algorithm does reduce the bases further than LLL. Looking at the conditions to be $\delta$-LLL reduced and $\omega-$Jacobi reduced, we can find for example two bases that are LLL reduced but not Jacobi reduced and will miss the shortest vector.

@@

Although the Jacobi method might have practical applications in signal processing, it is not appropriate for cryptographic applications on random lattices or higher dimensional bases.
